Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1]}

### 2018-01-12 paper/paper.tex
Bettererer

\newcommand{\citet}[1]{\citeauthor{#1} \shortcite{#1}}
\newcommand{\citep}[1]{\cite{#1}}

@@

\begin{abstract}
    Modern subgraph isomorphism solvers use value-ordering heuristics to direct a constraint
    programming style search. Although these heuristics are far from perfect, discrepancy-based
    searches are too expensive to be effective in this setting.  However, we show that a random
    value-ordering heuristic with restarts beats conventional search with the best-known tailored
    heuristic. This motivates the introduction of a slightly-random value-ordering heuristic with
    similar behaviour to the tailored heuristic: when combined with restarts, this new search
    strategy is over ?? times more effective on satisfiable instances, and does not
    significantly decrease performance on unsatisfiable instances. Finally, we show that this
    strategy also improves two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to find a copy of a small ``pattern'' graph inside a larger
``target'' graph.  The current single strongest subgraph isomorphism solver uses ``highest degree
first'' as a value-ordering heuristic to direct a constraint programming style search
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early on in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and restarts (which are
extremely beneficial). To explore different parts of the search space when restarting, we introduce
a new value-ordering heuristic, which adds a small amount of randomness to the ``highest degree
first'' heuristic.  The effects of the changes we propose are shown in
\cref{figure:scatter-by-family}: we can make a state-of-the-art algorithm perform much better on a
large number of satisfiable instances, whilst performing worse only rarely on satisfiable instances,
and never on unsatisfiable instances.  Further, these properties are \emph{not} shared by
discrepancy searches (\cref{figure:scatter-dds}) or by random value-ordering with restarts
(\cref{figure:scatter-random}). Finally, we show that this strategy is also effective in an
optimisation setting, producing clear benefits in two maximum common induced subgraph algorithms.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent. The \emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$,
is the set of vertices adjacent to $v$, and the \emph{degree} of $v$, $\deg_G(v)$, is the
cardinality of its neighbourhood; when the graph is clear from the context, we omit the $G$
subscript. The \emph{neighbourhood degree sequence} of $v$, $\nds_G(v)$, is the sequence of degrees
of the neighbours of $v$ in descending order, and we denote by $\preceq$ the inclusion relation of
\citet{DBLP:journals/constraints/ZampelliDS10}. As per \citet{DBLP:conf/aaai/HoffmannMR17},
$G^{n,\ell}$ denotes the graph with vertex set $\vertexset(G)$, and with edges between vertices $v$
and $v$ if there are at least $n$ simple paths of length exactly $\ell$ between vertices $v$ and $w$
in $G$.

@@

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we will leave it intact, so as not to affect the performance of the
algorithm on unsatisfiable instances.  For value ordering, which occurs on
\siplineref{valueordering} of \cref{algorithm:sip}, we will initially select vertices from highest
degree to lowest degree; we return to value ordering in the second half of this paper.

?? note that variable ordering affects unsat, whereas value ordering affects only sat.

@@

We use the dataset introduced by \citet{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers.
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families because some earlier solvers find many of these instances extremely hard. We
would like to have a larger number of satisfiable instances in our test set, and so we include all
pattern / target pairs. This gives a total of 14,621 instances (rather than the original 5,725). At
least ?? of these instances are known to be satisfiable, and at least ?? are unsatisfiable.

@@

We begin by verifying that our implementation of the basic \cref{algorithm:sip} is competitive. In
\cref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \cref{algorithm:sip} is derived
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \citep{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\citep{DBLP:journals/pami/CordellaFSV04}. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout; the horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances solved by the two algorithms could be completely different).

@@

\Cref{figure:scatter-dds} shows the effects of adding \citet{DBLP:conf/ijcai/Walsh97}'s
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

@@

\Cref{figure:restarts} shows that either biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.

@@

?? Benchmark results on the same families of instances are shown in
\cref{figure:mcsplit,figure:kdown,figure:mcs}. Clearly beneficial. Note that on McSplit we're doing
extra work and this shows, but it's still worthwhile.

>>>>

\newcommand{\citet}[1]{\citeauthor{#1} \shortcite{#1}}
\newcommand{\citep}[1]{\cite{#1}}
\newcommand{\citepos}[1]{\citeauthor{#1}'s \shortcite{#1}}

@@

\begin{abstract}
    Modern subgraph isomorphism solvers use degree-based heuristics to direct a constraint
    programming style search. We show that discrepancy-based searches are too expensive to be
    effective in offsetting mistakes made by value-ordering heuristic in this setting.  We then
    demonstrate that a random heuristic with restarts beats conventional search with the best-known
    tailored heuristic. This motivates the introduction of a slightly-random heuristic with similar
    behaviour to the tailored heuristic: when combined with restarts, this new search strategy is
    over a hundred times more effective on satisfiable instances, and does not affect performance on
    unsatisfiable instances. Finally, we show that the same strategy also improves two maximum
    common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to find a copy of a small ``pattern'' graph inside a larger
``target'' graph.  The current single strongest subgraph isomorphism solver uses ``highest degree
first'' as a value-ordering heuristic to direct a constraint programming style search
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early on in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and restarts (which are
extremely beneficial). To explore different parts of the search space when restarting, we introduce
a new value-ordering heuristic, which adds a small amount of randomness to the ``highest degree
first'' heuristic.  The effects of these changes are shown in
\cref{figure:scatter-by-family}: we can make a state-of-the-art algorithm perform much better on a
large number of satisfiable instances, whilst performing worse only rarely on satisfiable instances,
and never on unsatisfiable instances.  Further, these properties are \emph{not} shared by
discrepancy searches (\cref{figure:scatter-dds}) or by random value-ordering with restarts
(\cref{figure:scatter-random}). Finally, we show that this strategy is also effective in an
optimisation setting, producing benefits in two maximum common induced subgraph algorithms.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood; when the graph is
clear from the context, we omit the $G$ subscript. The \emph{neighbourhood degree sequence} of $v$,
$\nds_G(v)$, is the sequence of degrees of the neighbours of $v$ in descending order, and we denote
by $\preceq$ the inclusion relation of \citet{DBLP:journals/constraints/ZampelliDS10}. As per
\citet{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with vertex set $\vertexset(G)$,
and with edges between vertices $v$ and $v$ if there are at least $n$ simple paths of length exactly
$\ell$ between vertices $v$ and $w$ in $G$.

@@

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we will leave it intact, so as not to affect the performance of the
algorithm on unsatisfiable instances.  For value ordering, which occurs on
\siplineref{valueordering} of \cref{algorithm:sip}, we will initially select vertices from highest
degree to lowest degree; we return to value ordering in the second half of this paper. Note that the
variable-ordering heuristic affects both unsatisfiable and satisfiable instances, whilst the
value-ordering heuristic is relevant only for satisfiable instances.

@@

We use the dataset introduced by \citet{DBLP:conf/lion/KotthoffMS16} for evaluation. This dataset
brings together a range of randomly-generated and application instance families from earlier papers.
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families because some earlier solvers find many of these instances extremely hard. We
would like to have a larger number of satisfiable instances in our test set, and so we include all
pattern / target pairs. This gives a total of 14,621 instances (rather than the original 5,725). At
least \input{gen-n-sat.tex}\unskip\ of these instances are known to be satisfiable, and at least
\input{gen-n-unsat.tex}\unskip\ are unsatisfiable.

@@

We begin by verifying that our implementation of the basic \cref{algorithm:sip} is competitive. In
\cref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \cref{algorithm:sip} is derived
\citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \citep{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\citep{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout; the horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances solved by the two algorithms could be completely different).

@@

\Cref{figure:scatter-dds} shows the effects of adding \citepos{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

@@

\Cref{figure:restarts} shows that the biased heuristic together with restarts is better still---that
is, if we are introducing restarts, then it is better to add a small amount of randomness to a
tailored deterministic heuristic than it is to throw away the heuristic altogether.

@@

?? Benchmark results on the same families of instances are shown in
\cref{figure:kdown,figure:mcsplit,figure:mcs}. Clearly beneficial. Note that on McSplit we're doing
extra work and this shows, but it's still worthwhile.
